Thesis Proposal

### 2022-03-01 proposal/proposal.tex
add planning section

\title{Thesis Proposal}
\author{Hossein Naderi}
\maketitle

@@

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a linked list simultaneously. One solution is to use locks; whenever a process wants to do an update query on a data structure, it locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow parallelism since only one process can make progress in one moment. Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.

\paragraph{}
The question that may arise is, ``What properties matter for a lock-free data structure?''.
Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A safety condition tells us the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.

The standard safety condition is called linearizability, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure 1 is an example of execution on a linearizable queue. The arrow shows time, and rectangles show the time between invocation and termination of an operation. Since ENQ(A) and ENQ(B) are concurrent ENQ(B) may have taken effect before ENQ(A). Execution in figure 2 is not consistent since A has been enqueued before B, so it has to be dequeued first.

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/good}
  \caption{An example of a linearizable execution. Both could take effect first since ENQ(A), and ENQ(B) are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[width=3in]{pics/bad}
  \caption{An example of an execution that is not linearizable. Since ENQ(A) has completed before ENQ(B) is invoked the DEQ() should return A.}
\end{figure}

An algorithm is \textit{wait-free} if each operation terminates after a finite number of steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free. From now on, we are going to focus on wait freedom.

@@

In the following paragraphs, we look at previous lock-free queues.
Michael \& Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. It is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly~linked~list. Head points to the first not dequeued node in the linked list and tail points to the last element in the queue. To insert a node to the linked list, they use atomic primitive operations like LL/SC or CAS. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the worst-case number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can $\Omega(p)$ take steps.

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/msqueue}
  \caption{MS-queue structure, Enqueue and Dequeue operations}
\end{figure}

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea. The elimination mechanism benefits the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing operations such as consecutive enqueues and dequeues when the queue is initially empty in the elimination array to exchange values. Their algorithm makes it possible for failed aged operations to complement. Empirical evaluation of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes’ enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Operations in a basket try to find their order in the basket one by one by using CAS operations. However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/baskets}
  \caption{Baskets queue idea}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues. They use a doubly linked list and do fewer compare and swap operations than MS-queue. But as before, the worst case is when there are $p$ concurrent enqueues which have to be enqueued one by one. The amortized worst-case complexity is still $\Omega(p)$ CASes.

@@

Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. Global head and tail point to arrays and content of arrays are marked if dequeued and written by CAS operations. Their data structure is lock-free, but it updates lazily. Threads have a cache view on the queue and update it if it gets old.

@@

%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
In the worst-case step complexity of all the papers discussed above, there is a $p$ term which comes from the case all $p$ processes try to do an enqueue at the same time. We are focusing on seeing if we can do this in sublinear in terms of $p$ or not.

@@

\section{Our contribution}
In this work, we are trying to design a queue with $\log p^2 +Q$ steps per operation, outperforming queues built by LinkedList. Our idea is similar to the Jayanti MESD queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use CAS operations with big words. The new work about our idea is that even if there are $p$ processes trying to enqueue simultaneously, each enqueue takes $\log p^2 +Q$. Jayanti~\cite{DBLP:conf/podc/Jayanti98a} proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions. He also introduced~\cite{DBLP:conf/podc/ChandraJT98} a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses. We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ CAS per operation. We use the block tree as a stepping stone towards our queue algorithm.
\paragraph{}
A block tree is a tournament tree shared among $p$ processes. Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate up to the tree's root. An ordering is stored in each node of the tree of operations propagated to the node. All processes agree on the sequence stored in the root (Linearization ordering). 
\begin{figure}
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{Each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root which the final ordering is stored.}  
\end{figure}

The goal is to make each propagate step ensure that the operations are propagated up to the root and fast. And then use the linearization ordering to answer the dequeue operations.

In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using CAS operations. The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$. Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$) which has appended the talked operations. To make CAS fast, we do not append operations explicitly, but only remember their number. This allows us to CAS fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure.

\begin{figure}
\begin{center}
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{Operations after $||$ are new.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{New operations are added to the parent node.}
\end{subfigure}
\caption{\label{fig:ucexample} Propagate Step, operations in children after || are new.}
\end{center}
\end{figure}

We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization. Get(i) finds the block containing $i$th in the root and then finds its subblock recursively to reach a leaf. Index() is somewhat similar but more complicated from bottom to top; the main challenge is that each level of the tree should take polylogarithmic steps with respect to $p$. Having methods Get(i), Index(op), we can have the linearization shared among the processes. We will implement Enqueue and Dequeue using our block tree to have a queue. An Enqueue can be done using just an Append operation in the block tree. Dequeue() consists of an Append, Index, compute the corresponding response, and return it using Get().

However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing $i$th operation in the root level. We make it fast by storing prefix sums but still, it is proportional to the number of all operations. We come with a garbage collection only for the root level to reach $\log Q$. The main idea is to make use of the fast split of persistent red-black trees.

@@

>>>>

\title{Thesis Proposal}
\author{Hossein Naderi \and Eric Ruppert}
\maketitle

@@

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures; for example, consider two processes trying to insert some values to a linked list simultaneously. One solution is to use locks; whenever a process wants to do an update query on a data structure, it locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only one process holding the lock can make progress in the moment. Our approach is to create a lock-free data structure that developers can safely use without the disadvantages of locks.

The question that may arise is, ``What properties matter for a lock-free data structure?''.
Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A safety condition tells us that the data structure does not return wrong responses, and a progress condition indicates that operations eventually terminate.

The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} is an example of execution on a linearizable empty queue. The arrow shows time, and rectangles show the time between invocation and termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not have taken effect before \texttt{Enqueue(A)}. Execution in figure \ref{fig::badexample} is not consistent since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/good}
  \caption{\label{fig::goodexample}An example of a linearizable execution. Both could take effect first since \texttt{Enqueue(A)}, and \texttt{Enqueue(B)} are concurrent.}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution that is not linearizable. Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A}.}
\end{figure}

An algorithm is \textit{wait-free} if each operation terminates after a finite number of steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free.
%From now on, we are going to focus on wait freedom.

@@

In the following paragraphs, we look at previous lock-free queues.
Michael \& Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. It is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly~linked~list. Head points to the first not dequeued node in the linked list and tail points to the last element in the queue. To insert a node to the linked list, they use atomic primitive operations like LL/SC or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the worst-\texttt{CAS}e number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.5]{pics/msqueue}
  \caption{MS-queue structure, Enqueue and Dequeue operations}
\end{figure}

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination idea. The elimination mechanism benefits the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing operations such as consecutive enqueues and dequeues  in the elimination array to exchange values when the queue is empty. Their algorithm makes it possible for failed aged operations to complement. Empirical evaluation of their work is better than MS-queue, but the worst \texttt{CAS}e is still the same; in \texttt{CAS}e there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried to make the MS-queue more parallel by introducing the Baskets Queue. Their idea is to allow more parallelism by treating the simultaneous enqueue operations as a basket. Each basket has a time interval in which all its nodes’ enqueue operations overlap. Since the operations in a basket are concurrent, we can order them in any way. Operations in a basket try to find their order in the basket one by one by using \texttt{CAS} operations. However, like the previous algorithms, if there are still $p$ concurrent enqueue operations in a basket, the amortized complexity remains $\Omega(p)$.

\begin{figure}[hbt]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea}
\end{figure}

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08} presented an Optimistic Approach to Lock-Free FIFO Queues. They use a doubly linked list and do fewer compare and swap operations than MS-queue. But as before, the worst \texttt{CAS}e is when there are $p$ concurrent enqueues which have to be enqueued one by one. The amortized worst-\texttt{CAS}e complexity is still $\Omega(p)$ \texttt{CAS}es.

@@

Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. Global head and tail point to arrays and content of arrays are marked if dequeued and written by \texttt{CAS} operations. Their data structure is lock-free, but it updates lazily. Threads have a cache view on the queue and update it if it gets old.

@@

%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
In the worst-\texttt{CAS}e step complexity of all the papers discussed above, there is a $p$ term which comes from the \texttt{CAS}e all $p$ processes try to do an enqueue at the same time. We are focusing to see if we can do this in sublinear steps in terms of $p$ or not.

@@

\section{Our contribution}
In this work, we design a queue with $\log^2 p +Q$ steps per operation, outperforming queues built by the LinkedList approach. Our idea is similar to Jayanti 's Multi Enqueuer-Single Dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words. The new work about our idea is that even if there are $p$ processes trying to enqueue or dequeue simultaneously, each operation takes $\log^2 p +Q$. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-\texttt{CAS}e shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced~ a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. We introduce a data structure that makes processes agree on the linearization ordering using $\textsc{O}(\log p)$ \texttt{CAS} per operation called \textit{Block tree}. Then we use the block tree as a stepping stone towards our queue algorithm.
A block tree is a tournament tree shared among $p$ processes (See Figure~\ref{fig::blocktree}). Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate up to the tree's root. An ordering is stored in each node of the tree of operations propagated to the node. All processes agree on the sequence stored in the root (Linearization ordering). 
\begin{figure}
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{ّ\label{fig::blocktree}In the block tree each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root which the final ordering is stored.}  
\end{figure}

The goal here is to ensure in each propagate step, the new operations are propagated up to the parent in $\log p$ steps (See Figure \ref{fig::propagstep}). And then use the linearization ordering to answer the dequeue operations.

In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using \texttt{CAS} operations. The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in the first \textsc{Refresh}($n$) are guaranteed to be in $n$. Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations. Instead of storing operations explicitly in the nodes, we only keep track of the number of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from left and right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.

\begin{figure}
\begin{center}
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{\label{fig::propagstep}Operations after $||$ are new.}
\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}}
  \caption{New operations are added to the parent node.}
\end{subfigure}
\caption{\label{fig:ucexample} Propagate Step, operations in children after || are new.}
\end{center}
\end{figure}

We also implement methods Get(i), Index(op) to get the $i$th propagated operation and compute the order of a propagated operation in the linearization. Get(i) finds the block containing the $i$th in the root and then finds its sub-block recursively to reach a leaf. Index() is somewhat similar but more complicated from bottom to top; the main challenge is that each level of the tree should take polylogarithmic steps with respect to $p$. Having methods Get(i), Index(op), we can have the linearization shared among the processes. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} can be done using just an \texttt{Append()} operation with input argument \texttt{e} in the block tree. \texttt{Dequeue()} consists of an \texttt{Append()}, \texttt{Index()}, compute the corresponding response, and return it using \texttt{Get()}.

\texttt{Get()} and \texttt{Index()} search among blocks in each level of the tree to find the containing sub-block or super-block of the given operation. We use prefix sum for binary search and some extra constant information that we do not discuss here.

However, the algorithm works as a queue, but the Get(i)  may take a long time since it has to find the block containing the $i$th operation at the root level. We come with a garbage collection only for the root level to reach $\log Q$. The main idea is to make use of the fast split of persistent red-black trees. So our algorithm works in $O(\log^2 p +Q)$ steps.

@@

\section{Planning}
The plan is to finish the writing of the thesis by Aug 2021. The thesis will contain:
\begin{itemize}
  \item More detailed discussion of previous work
  \item Description of the algorithm, layer by layer
  \item Pseudocode of the algorithm in two parts; regarding memory management
  \item Proof of linearizability and time analysis
\end{itemize}

We expect to submit a paper of our work to \textit{36th Int-| Symposium on Distributed Computing
October 25-27, 2022, Augusta, Georgia, USA} by May 13.
